HiC Contact Map Comaprison Using Graphlet Approach

### 2018-03-26 papers/manuscripts/manuscript.tex
lost track
modified manuscript

We divided the task of graphlet comparison into two parts: first we compare
graphlets from each contact map in normal cell lines (MIT) with the 
same contact maps in the other three Leukemic cells. Second we compare 
contact maps in a similar way but this time only between leukemic cells.
In the former case, the null hypothesis is that there is no difference
between contact maps of normal cells and leukemic cells and in the latter
case, the null hypotheis is that there is no difference between 
different leukemic cells.

We consider two measures of \textit{difference} when comparing contact
map graphlets across cell lines. 
The first measure is \textit{signature distance vectors} between
each contact map of two cell lines. 
Let  $\mathbf{S}^A_{ij}$  and $\mathbf{S}^B_{ij}$ be
signature matrices corresponding  to cells A and B
respectively, then we can caclulate the 
distance between each
loci (row) $l$ in $S^A$ and the the same loci in $S^B$
using the following formula from \cite{prvzulj2007biological}:

\begin{equation}
    D(\mathbf{S}^A_{ij}, \mathbf{S}^B_{ij}) = 
    \frac{1}{73}\sqrt{\sum_{o=0}^{72}{\mathbf{D}_{.o}^2}}
    \label{eq:distance_total}
\end{equation}
where elements of $\mathbf{D}$ can be calculated as follows:
\begin{equation}
    D_{lo} = w_o \times 
        \frac{log(S_{ijlo}^A+1) - log(S_{ijlo}^B+1)}
             {log(max(S_{ijlo}^A, S_{ijlo}^B) + 2)}
    \label{eq:distance_single}
\end{equation}

@@

The second measure of comparison that we use captures how 
similar two orbits are in terms of their count 
frequencies across loci between two contact maps. 
Each column in $S_{ij}$ can provide information
regarding the \textit{frequency distribution} of orbits throughout
the contact map $C_{ij}$. 
We can find how similar these distributions are to each other using
correlation measures.
These correlations can be calculated as the following:
\begin{equation}
    Corr(\mathbf{S}^A_{ij}, \mathbf{S}^B{ij}) = \mathbf{c}_{ij}
    \label{eq:correlation}
\end{equation}
where each element of $\mathbf{c}$ is calculated as below:
\begin{equation}
    c_{ijo} = Corr(\mathbf{S}^A_{ij.o}, \mathbf{S}^B_{ij.o})
    \label{eq:correlation}
\end{equation}

For each contact map,  calculated
similarity between orbit distributions using Pearson's r 
correlation, which is computationally efficient.
However, pearson's r might not be able to capture
non-functional relationships between distributions. As a result, we
also used Maximal Information Coefficient (MIC) 
\cite{reshef2011detecting} in order to compare
correlations. MIC calculates mutual information (MI) between two
distributions, but utilizes dynamic programming in order adjust
bin sizes and numbers in order to achieve highest MI.
MIC values between two variables fall between 0 and 1,
with 0 meaning the two variables are completely independent
and 1 meaning one is dependant on the other.
We used both Pearson's r and MI in order to compare orbit
frequencies. Although results from both approaches were more
or less consistent, MIC showed higher robustness than Pearson's 
r method.

\subsection{Hypotheses}

With regard to the two aforementioned comparison methods, we define
our two null hypotheses as the following:

\begin{enumerate}
    \item Average signature vector distance between each 
        contact map of MIT cells and
        their counterparts in the three leukemic cells
        are equal to $0$. In mathematical terms:
        $\forall i,j \in \{1, 2 ..., 24\}$
        \begin{align}
            \begin{split}
                H_0&: \frac{1}{|V_{ij}|}\sum_{l=0}^{|V_{ij}|}{d_{ijl}} = 0 \\
                H_1&: \frac{1}{|V_{ij}|}\sum_{l=0}^{|V_{ij}|}{d_{ijl}} > 0
            \end{split}
            \label{eq:h1}
        \end{align}

        where 
        $\mathbf{d}_{ij} = D(S^{MIT}_{ij}, S^L_{ij})$ and $L \in \{ALL, RL, CALL4\}$.

        \item 
            \begin{enumerate}
                \item Average orbit correlations within a contact map of MIT cell, 
                    and its counterpart in leukemic cells is equal to 1
                    for each contact map.
                Again in mathematical terms:
                $\forall i,j \in \{1, 2 ..., 24\}$
                \begin{align}
                    \begin{split}
                        H_0&: \frac{1}{73}\sum_{o=0}^{72}{c_{ijo}} = 1 \\
                        H_1&: \frac{1}{73}\sum_{o=0}^{72}{c_{ijo}} < 1
                    \end{split}
                    \label{eq:h2b}
                \end{align}
                        where $\mathbf{c}_{ij} = Corr(S^{MIT}_{ij}, S^L_{ij})$, $L$ 
                        is the same as above and $N$ 
                        is the total number of contact maps which 
                        is equal to $275$ in our case.

                \item Average orbit correlations  of MIT cell
                    and leukemic cells across contact maps is equal to 1
                    for each orbit.
                Again in mathematical terms:

                        \smallskip

                    $\forall o \in \{0, 1 ... 72\}$
                \begin{align}
                    \begin{split}
                        H_0&: \frac{1}{N}\sum_{i=1}^{23}{\sum_{j=i}^{23}{c_{ijo}}} = 1 \\
                        H_1&: \frac{1}{N}\sum_{i=1}^{23}{\sum_{j=i}^{23}{c_{ijo}}} < 1
                    \end{split}
                    \label{eq:h2a}
                \end{align}
                        $L$ and $\mathbf{c}_{ij}$ are the same as above.

            \end{enumerate}
\end{enumerate}
\section{results}

\subsection{Hypothesis 1: Orbit Distance Comparison}
\begin{figure*}[t]
    \centering
    \begin{subfigure}[b]{\textwidth}
        \includegraphics[width=\textwidth]{figures/orbit-distances_all.png}
        \caption{}
        \label{fig:orbit-distances_all}
    \end{subfigure}

    \begin{subfigure}[b]{\textwidth}
        \includegraphics[width=\textwidth]{figures/orbit-distances_intra.png}
        \caption{}
        \label{fig:orbit-distances_intra}
    \end{subfigure}
    \caption{   Pair-wise graphlet signature difference for
                (a) all 276 contact maps.
                (b) only the 23 intra-chromosomal contact maps.
                Each point on a graph is the result of averaging all
                the distances acroos all loci of that contact map.
                As can be viewd, although inter-chromosomal graphlets 
                do not show significant differences, RL's inter-chromosomal
                graphlets seem to be closer to MIT's. Also in some contact
                maps ALL graphlets show greater distance from MIT graphlets
                than the other two cell lines.
             }
    \label{fig:orbit-distances}
\end{figure*}
Result of pair-wise contact map graphlet distances can is
demonstrated in Figure \ref{fig:orbit-distances}.
Each point on the graph is the
average distance between signature vector of two same loci 
in the two cell lines as in Equation \ref{eq:h1}.
in order to test hypothesis \ref{eq:h1} a one sided t-test
was conducted. The resulting p-values showed highly significant differeces
between graphlets extracted from all contact maps. Refer to supplementary
material for results of all hypotheis tests.

\subsection{Hypothesis 2: Orbit distribution comparison}
We caclulated pair-wise Pearson's r correlation for each orbit in each of the 276
contact maps from MIT data and ALL, RL and CALL4 data separately. 
Figure \ref{fig:contact_maps_correlations}
shows average orbit correlations across all contact maps
(average along the red \textit{vertical} arrow in figure 
\ref{graphlet_correlation_schema}), while
figure \ref{fig:orbits_correlations} 
demonstrates average correlations across all 72 orbits within each  contact map
(average along the red \textit{horizontal} arrows in figure 
\ref{graphlet_correlation_schema}).
It is worth mentioning that \textit{interchromosomal} thresholded contact maps 
represent
a biparticde graph with the loci from each chromosome on one side. Due to this
bipartide nature of the graphs in inter-chromosomal maps,
count of certain orbits is always 0, resulting in
a correlation values of 0 for them as well.
We ignored these values  when we calculated averages across orbits 
in figure \ref{fig:contact_maps_correlations_all} since they
would result in a bias towards zero in averages. You can see the bias in 
figure \ref{fig:orbits_correlations_all} where average correlations of obits
3, 9, 10-14, 20-34, 39-48, and 51-72 are close to zero. In fact all correlations
corresponding to these orbits are 0 except for the ones between the same 
chromosomes which are illustrated in Figure 
\ref{fig:orbits_correlations_intra}.

>>>>

%We divided the task of graphlet comparison into two parts: first we compare
%graphlets from each contact map in normal cell lines (MIT) with the 
%same contact maps in the other three Leukemic cells. Second we compare 
%contact maps in a similar way but this time only between leukemic cells.
%In the former case, the null hypothesis is that there is no difference
%between contact maps of normal cells and leukemic cells and in the latter
%case, the null hypotheis is that there is no difference between 
%different leukemic cells.

We consider two measures of \textit{difference} when comparing contact
map graphlets across cell lines. 
The first measure is \textit{signature distance vectors} between
each contact map of two cell lines. 
For a pair of cells A and B, let 
$\mathbf{S}^A_{ij}$  and $\mathbf{S}^B_{ij}$ be their
signature matrices. The \textit{signature distance} of
contact map $\mathbf{C}_{i,j}$ between A and B is denoted
by $\mathbf{d}^{A,B}_{ij}$. $\mathbf{d}^{A,B}_{ij}$ 
is a vector of size $|V_{i,j}|$
and its elements $d^{A,B}_{i,j,l}$ are
calculated using the following formula from \cite{prvzulj2007biological}:

\begin{equation}
    d^{A,B}_{i,j,l} = 
    \frac{1}{73}\sqrt{\sum_{o=0}^{72}{t_{lo}^2}}
    \label{eq:distance_total}
\end{equation}

where elements of $t_{i,j,l,o}$ is the
distance between each
loci (row) $l$ in $\mathbf{S}^A$ and the the same loci in 
$\mathbf{S}^B$ for
orbit $o$ as is calculated as below:

\begin{equation}
    t_{lo} = w_o \times 
    \frac{log(S_{ijlo}^A+1) - log(S_{ijlo}^B+1)}
    {log(max(S_{ijlo}^A, S_{ijlo}^B) + 2)}
    \label{eq:distance_single}
\end{equation}

@@

The second measure of comparison that we use captures how 
similar two orbits are in terms of their count 
frequencies across loci between two contact maps. 
Each column in $S_{ij}$ can provide information
regarding the \textit{frequency distribution} of orbits throughout
the contact map $C_{ij}$. 
We can find how similar these distributions are to each other using
correlation measures.
These correlations are denoted by $\mathbf{m}^{A,B}_{i,j}$ and
can be calculate using
any plausible correlation measure. 
In this study, for each contact map,  we calculated
similarity between orbit distributions using Pearson's r 
correlation, which is computationally efficient.
However, pearson's r might not be able to capture
non-functional relationships between distributions. As a result, we
also used Maximal Information Coefficient (MIC) 
\cite{reshef2011detecting} in order to compare
correlations. MIC calculates mutual information (MI) between two
distributions, but utilizes dynamic programming in order adjust
bin sizes and numbers in order to achieve highest MI.
MIC values between two variables fall between 0 and 1,
with 0 meaning the two variables are completely independent
and 1 meaning one is dependant on the other.
We used both Pearson's r and MI in order to compare orbit
frequencies. Although results from both approaches were more
or less consistent, MIC showed higher robustness than Pearson's 
r method.

If MIC is used as correlation measure, each element of 
 $\mathbf{c}$ is calculated as below:
\begin{equation}
    m^{A,B}_{ijo} = MIC(\mathbf{S}^A_{ij.o}, \mathbf{S}^B_{ij.o})
    \label{eq:mic}
\end{equation}
Alternatively, if we use Pearson criterion we would have:
\begin{equation}
    m^{A,B}_{ijo} = Pearson(\mathbf{S}^A_{ij.o}, \mathbf{S}^B_{ij.o})
    \label{eq:pearson}
\end{equation}

\section{results}

Result of pair-wise contact map graphlet distances can is
illustrated in Figure \ref{fig:orbit-distances}.
Each point on the graph is the
average of the graphlet distance vector 
($\bar{\mathbf{d}}^{A,B}_{i,j}$)
of the two cell lines specified in the legend.
 A one sided paired t-test
was conducted . The resulting p-values showed highly 
significant differeces from zero for all pairs of
cells. Detailed results of t-tests for each
contact map can be fount in supplementary materials.
For each contact map, each pair of cells are
order based on whether they are statistially
larger that the other or not. For example
for contact map $C_{i,j}$, the results 
of t-test is as follows:\\
\smallskip
{
    \small
zero  $<$  call4-mit  $<$     all-rl  $<$     mit-rl  $<$   call4-rl  $<$    all-mit  $=$  all-call4
}\\
\smallskip
which means that $\mathbf{d}^{\scriptscriptstyle CALL4, MIT}_{1,1}$ is statistically larger than 0,
but less than $\mathbf{d}^{\scriptscriptstyle ALL, RL}_{1,1}$ 
and so forth. We can also conclude that $\mathbf{d}^{\scriptscriptstyle ALL, CALL4}_{1,1}$ is not
statistically different from $\mathbf{d}^{\scriptscriptstyle ALL, MIT}_{1,1}$
between graphlets extracted from all contact maps. 
Refer to supplementary
material for results of all hypotheis tests.

\begin{figure*}[t]
    \centering
    \begin{subfigure}[b]{\textwidth}
        \includegraphics[width=\textwidth]{figures/orbit-distances_all.png}
        \caption{}
        \label{fig:orbit-distances_all}
    \end{subfigure}

    \begin{subfigure}[b]{\textwidth}
        \includegraphics[width=\textwidth]{figures/orbit-distances_intra.png}
        \caption{}
        \label{fig:orbit-distances_intra}
    \end{subfigure}
    \begin{subfigure}[b]{\textwidth}
        \includegraphics[width=\textwidth]{figures/orbit-distances_intra_separate.png}
        \caption{}
        \label{fig:orbit-distances_intra_separate}
    \end{subfigure}
    \caption{   Pair-wise graphlet signature difference for
                (a) all 276 contact maps.
                (b) only the 23 intra-chromosomal contact maps.
                (c) all 276 contact maps as well as corresponding
                standard errors.
                Each point on a graph is the result of averaging all
                the distances acroos all loci of that contact map.
                As can be viewd, although inter-chromosomal graphlets 
                do not show significant differences, RL's inter-chromosomal
                graphlets seem to be closer to MIT's. Also in some contact
                maps ALL graphlets show greater distance from MIT graphlets
                than the other two cell lines.
             }
    \label{fig:orbit-distances}
\end{figure*}

We caclulated pair-wise MIC for each orbit in each of the 276
contact maps from MIT data and ALL, RL and CALL4 data separately. 
Figure \ref{fig:contact_maps_correlations}
shows average orbit correlations across all contact maps
($\bar{\mathbf{c}}^{\scriptscriptstyle A,B}_{i,j}$: average along the red \textit{vertical} arrow in figure 
\ref{graphlet_correlation_schema}), while
figure \ref{fig:orbits_correlations} 
demonstrates average correlations across all 72 orbits within each  contact map
(average along the red \textit{horizontal} arrows in figure 
\ref{graphlet_correlation_schema}).
It is worth mentioning that \textit{interchromosomal} thresholded contact maps 
represent
a biparticde graph with the loci from each chromosome on one side. Due to this
bipartide nature of the graphs in inter-chromosomal maps,
count of certain orbits is always 0, resulting in
a correlation values of 0 for them as well.
We ignored these values  when we calculated averages across orbits 
in figure \ref{fig:contact_maps_correlations_all} since they
would result in a bias towards zero in averages. You can see the bias in 
figure \ref{fig:orbits_correlations_all} where average correlations of obits
3, 9, 10-14, 20-34, 39-48, and 51-72 are close to zero. In fact all correlations
corresponding to these orbits are 0 except for the ones between the same 
chromosomes which are illustrated in Figure 
\ref{fig:orbits_correlations_intra}.
